(preliminary title) Model compression for Random Forests using leaf clustering

### 2025-12-29 reports/paper/main.tex

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{floatrow}
%\usepackage{natbib}
\usepackage{doi}
\usepackage{graphicx}
\usepackage{subcaption}

@@

%% Abstract
\begin{abstract}	\noindent
The Random Forest algorithm has been very successful, and continue to be a powerful choice on low-power microcontrollers with small amounts of memory and compute.
Current size-optimized implementations of Random Forest tend to use hard majority voting, where each decision tree returns the most probable class, instead of the class proportions (probabilities).
We show that the hard majority voting rule can lead to a considerable drop in predictive performance on some datasets.
To find a better tradeoff between predictive performance and model size, we propose to use quantization and K-means clustering to adaptively select a small set of class proportions to store for the decision trees.
We evaluate the proposed method on 60 classification datasets from the OpenML CC18 benchmark suite.
Find that only a few class probabilities need to be stored to retain good performance,
with small increase in model size over hard majority voting.
When combined with 16-bit integer quantization of the input features, the method reached average model compression of XX, when staying within XXX percentage point performance drop.
The method is implemented in the open-source TinyML library emlearn, and demonstrated running on microcontrollers for Human Activity recognition tasks.

@@

@@

\cite{random_forest_breiman2001}

@@

TODO: illustration of tree structure
TODO: illustration of the storage, of decision nodes and leaves
Structure of a random forest.
Decision nodes, leaf nodes.
Voting rules. argmax / Hard majority.
Probability average. Soft.

Fully grown versus partially grown trees.
Fully grown, all leaves with end up with class proportions.
However when there are restrictions. Such as \hyperparam{max_depth}, \hyperparam{min_samples_leaf}, \hyperparam{min_samples_split} etc.
In resource constrained environments, this is the more realistic case.

@@

@@

As implemented in scikit-learn \cite{scikit-learn} RandomForestClassifier.

@@

Optimization techniques for decision trees and decision tree ensembles such as Random Forest has been studied extensively over the last decades, including in a low-power, low-compute setting.

Integer quantization of features.
FPGA/hardware related.
SIMD tree evaluation.

@@

Software tools for deploying RF to TinyML. Open-source

\section{Methods}
What we cover: Feature quantization. Leaf quantization. Leaf clustering.

OpenML CC18 datasets. \cite{oml-benchmarking-suites}
To avoid needing data imputation, the datasets with missing values were ignored.
TODO: list exact identifiers for that which was eliminated

Train a general model. Using scikit-learn. \cite{scikit-learn}
Using 10 trees ({n_estimators}). And {min_samples_leaf=0.01}.
NOTE: clustering will probably save more in percent, the more leaves are used.
Increases both with the number of trees, and the effective depth
- but total model sizes in excess of 100 kB is less relevant for the typical usecase

TODO: consider other values for {min_samples_leaf} ?
Either multiple, or tune on datasets.
It will affect leaf growth.

@@

Numerical variables are scaled using RobustScaler.
Categorical variables are encoded using OrdinalEncoder.

Apply compression strategy to trained model. Re-running predictions on test set.

Cross validation using stratified k-fold.

? Demonstrator of performance on one or few TinyML tasks/datasets.
HAR: PAMAP. UCI DSADS. OPPORTUNITY ? F1 scores. 
Comparisons with other frameworks. Leaf cluster (emlearn), m2cgen, micromlgen
Hardware architectures. ARM Cortex M0+ (no FPU). ESP32 (FPU). Alt: Cortex M4F
TODO: visualization. Panel over HW,dataset of scatterplots with model size and inference time
Need to implement feature computation also.

\footnote{The code for reproducing the experiments can be found at https://github.com/jonnor/leaf-quantization}

@@

% TODO: run experiments for all hyperparameters restricting tree depth
% But probably no need to include all figures in the paper
% scikit-learn RandomForest hyperparameters affecting tree depth and leaf nodes (number/split/composition)
% direct restrictors on leaves
% min_weight_fraction_leaf, min_samples_leaf, max_leaf_nodes
% indirect restrictions
% max_depth, min_samples_split, min_impurity_decrease, ccp_alpha

@@

FIXME: check which leaf size is used. Should be 32-bit floating point for.
And int16 features?

@@

TODO: results with leaf clustering (and quantization) enabled
Thesis: is pareto optimal in terms of predictive performance vs model size, when compared to majority and full soft voting
? how to show that this holds across all (or majority) of datasets

@@

Here we only considered the use of Random Forest for classification.
The models are also used for regression.
Explore whether leaf clustering could also be useful for regression.

@@

Martin Stensgård

>>>>

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage[strings]{underscore} % make sure _ in strings like \texttt works 
\usepackage{floatrow}
%\usepackage{natbib}
\usepackage{doi}
\usepackage{graphicx}
\usepackage{subcaption}

@@

%% Abstract
\begin{abstract}	\noindent
The Random Forest algorithm has been very successful, and continue to be a powerful choice on low-power embedded systems with small amounts of memory and compute.
Current size-optimized implementations of Random Forest inference tend to use hard majority voting, where each decision tree returns the most probable class, instead of the class proportions (probabilities).
We show that while compact, the hard majority voting rule can lead to a large drop in predictive performance on some datasets.
To enable a better tradeoff between predictive performance and model size,
we propose two methods to compress the model:
using leaf quantization with deduplication, or leaf clustering with K-means.
By evaluating using 60 classification datasets from the OpenML CC18 benchmark suite,
we find that only a few class probabilities need to be stored to retain good performance, with small increase in model size over hard majority voting.
When combined with 16-bit integer quantization of the input features, reducing the number of trees, the method reached average model compression of XXX, when staying within XXX percentage point performance drop of the uncompressed baseline.
The method is implemented in the open-source TinyML library emlearn,
and demonstrated running on microcontrollers for Human Activity recognition tasks.

@@

Importance of small models.
Microcontroller. Limited program storage (FLASH) and memory (SRAM).

@@

Random Forest model was originally introduced by \cite{random_forest_breiman2001}

Decision tree ensembles have several properties that make them attractive
for resource-constrained environments.
Fast inference
Easy to implement with integer-only arithmetic.
High predictive performance with minimal hyperparameter tuning.
Works well with small and medium sized datasets.

@@

Structure of a random forest.

Logically, a decision tree ensemble consists of a set of binary decision trees.
Each tree has decision nodes that evaluate a set of if-else conditions,
comparing one of the features of the input data to a learned threshold value. 
At the leaves of the trees are the classification predictions for that tree.

The leaves can either be a class index, in which case the ensemble prediction is done by counting the number of votes and outputting the majority.

Probability average. Soft.
% TODO: have some references to proportions, when it was introduced?

% TODO: illustration of logical tree structure

There are multiple ways of representing the decision tree ensemble at inference time.
% TODO: some references to other approaches
% As implemented in scikit-learn \cite{scikit-learn} RandomForestClassifier.

Here we use an approach consisting of two tables:
a table of decision nodes, plus a table of leaf nodes.

Each decision node consists of feature index, threshold value, left child index and right child index.
The left/right indices can either be an offset to a new decision node, or an offset into the table of leaf nodes (encoded as negative values).
This memory layout is illustrated in Figure XXX: FIXME.
% TODO: illustration of the storage, of decision nodes and leaves

Each entry in the leaf node table has XXXX class proportions.

@@

If there are no restrictions on tree depth,
all leaves end up only containing samples from a single class.
In this circumstance, storing only the class index (instead of class proportions) preserves all the information in the tree.
However, when there are restrictions on tree depth
(for example with scikit-learn hyperparameters \hyperparam{max_depth}, \hyperparam{min_samples_leaf}, \hyperparam{min_samples_split}, et.c.)
the leaves will containe a mix of class proportions.

In resource constrained environments, this is the more realistic case.

@@

@@

Optimization techniques for decision trees and decision tree ensembles (such as Random Forest) has been studied extensively over the last decades, including in a low-power and low-compute settings.

Integer quantization of features.
FPGA/hardware related. FIXME: references

SIMD tree evaluation.
TODO: cite QuickScorer/RAPIDSCORER, Fast Inference of Tree Ensembles on ARM Devices

@@

Software tools for deploying RF to TinyML. Open-source
% TODO: cite m2cgen, emlearn, etc

\section{Methods}

What we cover: Feature quantization. Leaf quantization. Leaf clustering.

\subsection{Comparison of leaf compression strategies on OpenML CC18}

A subset of datasets from the OpenML CC18 benchmarking suite\cite{oml-benchmarking-suites} were used as a basis for the evaluation. 
To avoid needing data imputation, the datasets with missing values were dropped,
and to avoid datasets which require substantial feature selection or engineering, datasets more than FIXME:XX features were dropped.
The selection removed FIXME:XXX datasets, leaving a total of FIXME:XXX datasets. The complete list can be found in the results.
% MAYBE: list exact identifiers that which was eliminated?

Numerical variables are scaled using RobustScaler, and categorical variables are encoded using OrdinalEncoder.

Train a general model. Using scikit-learn. \cite{scikit-learn}

Apply compression strategy to trained model. Re-running predictions on test set.
Cross validation using k-fold, with k=5.

@@

All experiments were done with int16 as feature representation,
which reduces the size of decision nodes considerably compared to float32, while still allowing a sufficient range to give equivalent results on most datasets/tasks.
% FIXME: references to existing papers
We also found that int16 features resulted in the same performance as float32 on the selected datasets (not shown).
The conversion from original feature data to 16-bit integers was done using robust min-max scaling between 0.01 percentile and 99.9 percentile.

An overview of the different model compression strategies can be seen in Table \ref{fig:leaf_uniqueness}. 
The Baseline configuration uses the default scikit-learn hyper-parameters (as of version FIXME), 100 trees (\texttt{n_estimators}) and no limit to tree depth (\texttt{min_samples_leaf=1}).
All other configurations used 10 trees (more representative for models for resource constrained targets), and a range of tree depths were considered by trying \texttt{min_samples_leaf} values between 1,2,4,8,16,32,64,128.
For Quantize we use between 1-8 bit quantization, and for ClusterQ8 consider 1,2,4,8,16,32 clusters per class.

%NOTE: clustering will probably save more in percent, the more leaves are used.
%Increases both with the number of trees, and the effective depth
%- but total model sizes in excess of 100 kB is less relevant for the typical usecase

\begin{table}[h]
\centering
\begin{tabular}{llllll}
\hline
\textbf{Strategy} & \textbf{Trees} & \textbf{Leaf representation} & \textbf{Deduplication} & \textbf{Clustering} & \textbf{Options} \\
\hline
Baseline & 100 & float32 & No & No  \\
Reduced & 10 & float32 & No & No \\
Majority & 10 & none & Yes & No \\
Deduplicate & 10 & float32 & Yes & No \\
Quantize & 10 & uint8 & Yes & No & 1-8 bit quantization \\
ClusterQ8 & 10 & uint8 & Yes & Yes & 1,2,4,8,16,32 clusters per class \\
% ClusterQ4 & 10 & 4+4 bits in uint8 & Yes & Yes \\
\hline
\end{tabular}
\caption{Experimental configurations}
\label{tab:models}
\end{table}

\subsection{Evaluation on hardware for Human Activity Recognition}

? Demonstrator of performance on one or few TinyML tasks/datasets.
HAR: PAMAP. UCI DSADS. OPPORTUNITY ?
F1 scores.

Feature process. TODO: decide. MiniRocket ?? Any pre-selection?

Comparisons with other frameworks. Leaf cluster (emlearn), m2cgen, micromlgen

Hardware architectures.
RP2040 ARM Cortex M0+ (no FPU). 
Nordic NRF52840 ARM Cortex M4F
ESP32 (FPU).

TODO: visualization. Panel over HW, dataset of scatterplots with model size and inference time

\footnote{The code for reproducing the experiments can be found at https://github.com/jonnor/leaf-quantization}
% FIXME: double check link before submission

@@

% TODO: run experiments for all hyperparameters restricting tree depth
% But probably no need to include all figures in the paper

@@

FIXME: Leaf size used. Should be 32-bit floating point for leaf probabilities.
And int16 features?

@@

TODO: results with leaf clustering (and quantization) enabled
Thesis: is Pareto optimal in terms of predictive performance vs model size, when compared to majority and full soft voting

@@

Here we only considered the use of leaf quantization and leaf clustering for classification, however we suspect that these methods would also be beneficial for  regression tasks.
The methods may also be useful for gradient-boosted trees (such as XGBoost, LightGBM, etc).

@@

We would like to thank Martin Stensgård for multiple insightful and productive discussions about low-level optimization of decision tree inference on microcontroller CPU arcitectures.

Soundsensing has in the period 2022-2025 received funding from Research Council of Norway as part of the project EARONEDGE\cite{earonedge_nfr}.
